---
title: "Differential Gene Expression of RPMI 37C 2hr vs 0min"
author: "Edward Wallace"
date: "2023-04-24"
output: 
  html_document:
    toc: true
---


# Summary

Differential Gene Expression of RPMI vs YPD in 37C 2hr samples using DESeq2.

The medium was RPMI + 10% FBS, abbreviated here to RPMI without subsequently mentioning the serum.

This analysis does:

- EDIT

It concludes that... EDIT

## Load count data from shared script load_count_data.Rmd

```{r call_load_count_data, child = "load_count_data.Rmd"}
```

# DESeq on a subset of data

```{r subset_RPMI37C2hr_vs_0min, fig.height=8, fig.width=8}
sample_sheet_RPMI37C2hr_vs_0min <-
  dplyr::filter(sample_sheet,
                MediumTempTime %in% c("YPD_25C_0", "RPMIS_37C_120")) %>%
  # rename "RPMI+" to "RPMI" to avoid character substitutions in DESeq2
  dplyr::mutate(Medium = forcats::fct_recode(Medium, RPMI = "RPMI+"),
                Time = factor(Time))

dds_RPMI37C2hr_vs_0min <- 
  DESeqDataSetFromMatrix(countData = counts_all %>%
                           dplyr::select(sample_sheet_RPMI37C2hr_vs_0min$Code) %>%
                           magrittr::set_rownames(counts_all$Gene),
                         colData = sample_sheet_RPMI37C2hr_vs_0min,
                         design = ~ Time) %>%
  DESeq()

dds_RPMI37C2hr_vs_0min
```

```{r results_RPMI37C2hr_vs_0min}
results(dds_RPMI37C2hr_vs_0min)

resultsNames(dds_RPMI37C2hr_vs_0min)
```

```{r deseq_df_RPMI37C2hr_vs_0min}
deseq_df_RPMI37C2hr_vs_0min <- 
  biobroom::tidy.DESeqDataSet(dds_RPMI37C2hr_vs_0min) %>%
  dplyr::filter(term == "Time_120_vs_0") %>%
  dplyr::left_join(gene_info_df,
                   by = "gene") %>% 
  dplyr::select(gene, 
                geneName, 
                baseMean, 
                log2FC = estimate, 
                stderror,
                padj = p.adjusted)
```

## List DEGs most UP in RPMI+ vs YPD

We cut off at 2x changes with 5% FDR here.


```{r DEGs_RPMI_up, fig.width = 6, fig.height = 3}
# Make data frames of DEGs up and down, arranged by log2FC
DEGdf_RPMI_up2x_FDR5 <- 
  dplyr::filter(deseq_df_RPMI37C2hr_vs_0min,
                log2FC > 1, padj < 0.05) %>%
  dplyr::arrange(desc(log2FC))
print(DEGdf_RPMI_up2x_FDR5, n = 20)
```

Print the top 20 *named* DEGs

```{r DEGs_RPMI_up_named}
DEGdf_RPMI_up2x_FDR5 %>%
  dplyr::filter(!is.na(geneName)) %>%
  print(n = 20)
```

## List DEGs most DOWN in RPMI+ vs YPD


```{r DEGs_RPMI_down, fig.width = 6, fig.height = 3}
# Make data frames of DEGs up and down, arranged by log2FC
DEGdf_RPMI_down2x_FDR5 <- 
  dplyr::filter(deseq_df_RPMI37C2hr_vs_0min,
                log2FC < -1, padj < 0.05) %>%
  dplyr::arrange(log2FC)
print(DEGdf_RPMI_down2x_FDR5, n = 20)
```

Print the top 20 *named* DEGs

```{r DEGs_RPMI_down_named}
DEGdf_RPMI_down2x_FDR5 %>%
  dplyr::filter(!is.na(geneName)) %>%
  print(n = 20)
```

## Volcano plot of DEGs

```{r volcano_RPMI37C2hr_vs_0min, fig.width = 6, fig.height = 3}
volcano_RPMI37C2hr_vs_0min <- 
  ggplot(data = deseq_df_RPMI37C2hr_vs_0min,
         aes(x = log2FC, y = -log10(padj)) ) +
  geom_point(size = 0.5, colour = "grey50") +
  geom_point(data = DEGdf_RPMI_up2x_FDR5,
             size = 1, colour = "darkblue") +
  geom_point(data = DEGdf_RPMI_down2x_FDR5,
             size = 1, colour = "darkred") +
  scale_y_continuous("-log10(p)",
                     limits = c(0,10), expand = c(0,0),
                     oob=scales::squish) +
  scale_x_continuous("log2 fold-change, RPMI+serum vs YPD, 37C 1hr",
                     breaks = -5:5,
                     limits = c(-4,4), expand = c(0.01,0.01),
                     oob = scales::squish) +
  coord_cartesian(clip = "off") 
volcano_RPMI37C2hr_vs_0min
```



## Write outputs to files

```{r write_DEGs}
deseq_df_RPMI37C2hr_vs_0min %>%
  dplyr::mutate(
    across(c(baseMean, log2FC, stderror), round, digits = 3)
  ) %>%
  dplyr::mutate(padj = signif(padj, digits = 4)) %>%
  write_tsv(file = here::here("results", "DEGs",
                              "deseq_df_RPMI37C2hr_vs_0min.txt"))

# DEGdf_RPMI_up2x_FDR5 %>%
#   dplyr::pull(gene) %>%
#   write_lines(file = here::here("results", "DEGs",
#                             "deglist_RPMI37C2hr_vs_0min_up2x_FDR5.txt"))
# 
# DEGdf_RPMI_down2x_FDR5 %>%
#   dplyr::pull(gene) %>%
#   write_lines(file = here::here("results", "DEGs",
#                             "deglist_RPMI37C2hr_vs_0min_down2x_FDR5.txt"))
```

