---
title: "Differential Gene Expression of 37C vs 25C in YPD samples"
author: "Edward Wallace"
date: "2022-09-05"
output: 
  html_document:
    toc: true
---


# Summary

Differential Gene Expression of 37C vs 25C in YPD samples at 1hr, using DESeq2

This analysis produces files:

- `deseq_df_37Cvs25C_YPD1hr.txt`
- `deglist_37Cvs25C_YPD1hr_up2x_FDR5.txt`
- `deglist_37Cvs25C_YPD1hr_down2x_FDR5.txt`

## Load count data from shared script load_count_data.Rmd

```{r call_load_count_data, child = "load_count_data.Rmd"}
```

# DESeq on a subset of data, only YPD at 1hr

```{r subset_37Cvs25C_YPD1hr, fig.height=8, fig.width=8}
sample_sheet_37Cvs25C_YPD1hr <-
  dplyr::filter(sample_sheet,
                Time == 60, Medium == "YPD")

dds_37Cvs25C_YPD1hr <- 
  DESeqDataSetFromMatrix(countData = counts_all %>%
                           dplyr::select(sample_sheet_37Cvs25C_YPD1hr$Code) %>%
                           magrittr::set_rownames(counts_all$Gene),
                         colData = sample_sheet_37Cvs25C_YPD1hr,
                         design = ~ Temp) %>%
  DESeq()

dds_37Cvs25C_YPD1hr
```
```{r results_37Cvs25C_YPD1hr}
results(dds_37Cvs25C_YPD1hr)

resultsNames(dds_37Cvs25C_YPD1hr)
```

```{r deseq_df_37Cvs25C_YPD1hr}
deseq_df_37Cvs25C_YPD1hr <- 
  biobroom::tidy.DESeqDataSet(dds_37Cvs25C_YPD1hr) %>%
  dplyr::filter(term == "Temp_37C_vs_25C") %>%
  dplyr::left_join(gene_info_df,
                   by = "gene") %>% 
  dplyr::select(gene, 
                geneName, 
                baseMean, 
                log2FC = estimate, 
                stderror,
                padj = p.adjusted)
```

## List DEGs most UP in 37C vs 25C

We cut off at 2x changes with 5% FDR here.


```{r DEGs_37C_up, fig.width = 6, fig.height = 3}
# Make data frames of DEGs up and down, arranged by log2FC
DEGdf_37C_up2x_FDR5 <- 
  dplyr::filter(deseq_df_37Cvs25C_YPD1hr,
                log2FC > 1, padj < 0.05) %>%
  dplyr::arrange(desc(log2FC))
print(DEGdf_37C_up2x_FDR5, n = 20)
```

Print the top 20 *named* DEGs

```{r DEGs_37C_up_named}
DEGdf_37C_up2x_FDR5 %>%
  dplyr::filter(!is.na(geneName)) %>%
  print(n = 20)
```


## List DEGs most DOWN in 37C vs 25C


```{r DEGs_37C_down, fig.width = 6, fig.height = 3}
# Make data frames of DEGs up and down, arranged by log2FC
DEGdf_37C_down2x_FDR5 <- 
  dplyr::filter(deseq_df_37Cvs25C_YPD1hr,
                log2FC < -1, padj < 0.05) %>%
  dplyr::arrange(log2FC)
print(DEGdf_37C_down2x_FDR5, n = 20)
```

Print the top 20 *named* DEGs

```{r DEGs_37C_down_named}
DEGdf_37C_down2x_FDR5 %>%
  dplyr::filter(!is.na(geneName)) %>%
  print(n = 20)
```

## Volcano plot of DEGs

```{r volcano_37Cvs25C_YPD1hr, fig.width = 6, fig.height = 3}
volcano_37Cvs25C_YPD1hr <- 
  ggplot(data = deseq_df_37Cvs25C_YPD1hr,
         aes(x = log2FC, y = -log10(padj)) ) +
  geom_point(size = 0.5, colour = "grey50") +
  geom_point(data = DEGdf_37C_up2x_FDR5,
             size = 1, colour = "darkblue") +
  geom_point(data = DEGdf_37C_down2x_FDR5,
             size = 1, colour = "darkred") +
  scale_y_continuous("-log10(p)",
                     limits = c(0,10), expand = c(0,0),
                     oob=scales::squish) +
  scale_x_continuous("log2 fold-change, 37C vs 25C, RPMI 1hr",
                     breaks = -5:5,
                     limits = c(-4,4), expand = c(0.01,0.01),
                     oob = scales::squish) +
  coord_cartesian(clip = "off") 
volcano_37Cvs25C_YPD1hr
```

The volcano plot shows a clear set of genes with log2FC less than 2 but adjusted p-value below 0.05.



## Write outputs to files

```{r write_DEGs}
deseq_df_37Cvs25C_YPD1hr %>%
  dplyr::mutate(
    across(c(baseMean, log2FC, stderror), round, digits = 3)
  ) %>%
  dplyr::mutate(padj = signif(padj, digits = 4)) %>%
  write_tsv(file = here::here("results", "DEGs",
                              "deseq_df_37Cvs25C_YPD1hr.txt"))

DEGdf_37C_up2x_FDR5 %>%
  dplyr::pull(gene) %>%
  write_lines(file = here::here("results", "DEGs",
                            "deglist_37Cvs25C_YPD1hr_up2x_FDR5.txt"))

DEGdf_37C_down2x_FDR5 %>%
  dplyr::pull(gene) %>%
  write_lines(file = here::here("results", "DEGs",
                            "deglist_37Cvs25C_YPD1hr_down2x_FDR5.txt"))
```

