---
title: "Quality control and PCA Template EDIT ME"
author: "Edward Wallace"
date: "13/06/2022"
output: 
  html_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(cowplot)
library(GGally)
library(here)
library(DESeq2)
library(biobroom)

theme_set(theme_cowplot(font_size = 12) +
    theme(# panel.grid.major = element_line(colour = "grey80", size = 0.5),
          strip.background = element_blank())
    )

# Count header columns from featurecounts
count_header_columns <- c("Geneid", "Chr", "Start",
                          "End", "Strand", "Length")

# unwanted string in filename; may need to be edited
unwanted_filename_string <- "_L001_R1_001_aln.bam"

```

# Summary - EDIT TO DESCRIBE YOUR DATA

Quality control of RNA-seq data after read processing.

This analysis does:

- Scatter plot of counts showing log-scale correlations
- PCA plot showing how samples are similar to one another

It concludes that... EDIT

# Load and organise raw count data

## Load the raw counts

```{r load_counts}
counts_raw <- 
  here::here("results", "counts", "counts.txt") %>%
  readr::read_tsv(comment = "#")
counts_raw
```

## Load the sample sheet

```{r load_samplesheet}
sample_sheet <- 
  here::here("input", "experiment",
             "samplesheet_RENAME.txt") %>%
  readr::read_tsv(comment = "#") 

sample_sheet 
```

## Organise the counts using the sample sheet

```{r organise_counts}
counts_all <- 
  counts_raw %>%
  dplyr::select(sample_sheet$Filename)  %>%
  dplyr::rename_with(stringr::str_remove,
                     pattern = unwanted_filename_string)%>%
  magrittr::set_rownames(counts_raw$Geneid)

counts_all
```

# Check reproducibility by paired scatter plot in log10-plus-1-space

_Note: This is set to `eval=FALSE` as it can run very slowly for large datasets._ 

```{r paired_scatter, fig.height=16, fig.width=16, eval = FALSE}
counts_all %>%
  magrittr::add(1) %>%
  log10() %>%
  GGally::ggpairs()
```

# Normalize counts by regularized log (rlog) with DESeq2

The [rlog function from DESeq2](https://rdrr.io/bioc/DESeq2/man/rlog.html) "transforms the count data to the log2 scale in a way which minimizes differences between samples for rows with small counts, and which normalizes with respect to library size."

## Construct DESeq dataset object

EDIT: the `design` formula may need to be edited.

```{r dds_counts_all}
dds_all <- DESeqDataSetFromMatrix(countData = counts_all,
                                  colData = sample_sheet,
                                  design = ~ .)
```

# Do the rlog transformation

```{r rlog_dds_all}
rlog_all <- rlog(dds_all)
head(assay(rlog_all))
```

# Check reproducibility by principal component analysis (PCA) of rlog

## Calculate PCs from the rlog output

```{r calculate_pca_rlog}
# calculate principal components of rlog, after extracting from the dataset
pca_rlog <- rlog_all %>%
  assay() %>%
  t() %>%
  prcomp()

# convert principal components to data frame
pcdf_rlog <- bind_cols(
  as_tibble(colData(rlog_all)),
  as_tibble(pca_rlog$x)
)

pcdf_rlog
```

## Calculate proportion of variance explained by each PC


```{r calculate_propvar}
propvar_rlog_df <- tibble(
  PC = seq.int(1L, ncol(pca_rlog$x) ),
  prop_var = pca_rlog$sdev^2/sum(pca_rlog$sdev^2)
)
```

## Plot PCA percentage variance explained to find how many PCs to pay attention to.

```{r plot_percentvar, fig.width = 3, fig.height = 2}
plot_percentvar_rlog <- 
  ggplot(data = propvar_rlog_df, 
         aes(x = PC, y = prop_var)) +
  geom_col(fill = "blue") +
  scale_x_continuous("principal component",
                     limits = c(0.4,10.6), 
                     # hack limits to include the bars and a little space.
                     breaks = 1L:10L,
                     expand = c(0,0)) + 
  scale_y_continuous("prop. of variance", expand = c(0,0))
plot_percentvar_rlog
```

## Plot PC1 vs PC2 with default values

EDIT: the colour and other aesthetics may need to be edited to highlight features of the data

```{r plot_PC12_default, fig.height=6, fig.width=6}
ggplot(data = pcdf_rlog,
       aes(colour = SampleID)
       ) +
  geom_point(aes(x = PC1, y = PC2)) 
```

## Plot PC1 vs PC2, 3, 4, also with default values

```{r plot_PC1234_text, fig.height=8, fig.width=8}
# This could be wraped in a function
pc_baseplot <- 
  ggplot(data = pcdf_rlog,
         aes(colour = SampleID, 
             label = SampleID)
  ) + 
  geom_hline(yintercept = 0, size = 0.5, colour = "grey80") +
  geom_vline(xintercept = 0, size = 0.5, colour = "grey80") +
  theme(legend.position = "none")

pcplot_12 <- pc_baseplot + geom_text(aes(x = PC1, y = PC2))
pcplot_13 <- pc_baseplot + geom_text(aes(x = PC1, y = PC3))
pcplot_14 <- pc_baseplot + geom_text(aes(x = PC1, y = PC4))
pcplot_23 <- pc_baseplot + geom_text(aes(x = PC2, y = PC3))
pcplot_24 <- pc_baseplot + geom_text(aes(x = PC2, y = PC4))
pcplot_34 <- pc_baseplot + geom_text(aes(x = PC3, y = PC4))

pcplot_legend <- get_legend(pcplot_12 + theme(legend.position = "right"))

plot_grid(pcplot_12, NULL, pcplot_legend, 
          pcplot_13, pcplot_23, NULL, 
          pcplot_14, pcplot_24, pcplot_34)
```
